### 2018-05-30 submission/manuscript.Rmd
complete changes to the discussion

Our observations support that the number of OTUs, error rate, and chimera prevalence depends on polymerase used and that chimera prevalence is highly correlated with the number of OTUs [Figure 1-5]. Our observations also suggest that undetected chimeras, not removed using standard bioinformatic approaches, cause many of these differences. In addition to the reported polymerase dependent differences, the bacterial community also varied by the number of cycles used. These differences are important because many diversity metrics rely on calculations (e.g. richness) that are dependent on the number of OTUs. Since the number of OTUs are dependent on error rate and chimera prevalence the diversity metrics can vary simply by changing the polymerase used. Based on these observations, metrics that measure diversity using calculations such as richness depend on the polymerase, but this may not be the case for all diversity metrics.

There were few differences that depend on polymerase for between sample diversity, as measured by the Bray-Curtis index. Our observations generally found no differences in the overall bacterial community composition for the number of cycles used. One possible reason for this outcome is that our study did not have enough power to detect differences due to low sample number in each group. Another reason is that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed polymerase dependent differences in chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in polymerases. When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. With differences in the number of OTUs and chimera prevalence depending on polymerase used, it might be easier to avoid specific DNA polymerase families altogether.

Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family, this is unlikely to be the only contributor. For example, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014]. Additionally, based on the material safety data sheet the differences between the two polymerases are not immediately apparent. Both polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences observed between the polymerases. We are unlikely to avoid adding polymerase dependent bias to 16S rRNA gene sequencing results, however, these differences may not be large enough to mask the actual biological signal.

The majority of polymerases we studied add small increases in the number of OTUs and chimera prevalence and may be masked by biological differences. The sequence error introduced by the polymerase is also small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014]. The choice of polymerase should be an important consideration in the creation of a 16S rRNA gene sequencing workflow because using different polymerases can add error and bias to the downstream observations. Although standardization of the workflow may partially solve this problem by introducing a consistent bias to all samples, it does come at a cost.

Methods can be standardized but they commonly contain bias that is reproducible and may miss important associations. Bias can be easily reproduced and can be found in every step of the 16S rRNA gene sequencing workflow [@Amir2017; @preservation_Luo_2016; @contamination_Salter2014; @taq_Gohl_2016]. This study shows that specific diversity metrics used to measure the microbial community consistently vary based on polymerase. Standardizing multiple workflows to one specific polymerase could be detrimental since some polymerases may work better in certain situations over others. Arguably, the degree of workflow standardization across studies and research group needs to be approached on a study by study basis and not every project needs to use the exact same approach. All aspects of the 16S rRNA gene sequencing workflow need to be customized for the specific microbial community that will be sampled. Although a diversity of approaches may make reproducibility and replicability more difficult it will help to avoid systematic biases from occuring due to widespread standardization of approaches.

>>>>

Our observations support that the number of OTUs, error rate, and chimera prevalence depends on polymerase used and that chimera prevalence is highly correlated with the number of OTUs [Figure 1-2 & 4-6]. Our observations also suggest that undetected chimeras, not removed using standard bioinformatic approaches, cause many of these differences. In addition to the reported polymerase dependent differences, the bacterial community also varied by the number of cycles used. These differences are important because many diversity metrics rely on calculations (e.g. richness) that are dependent on the number of OTUs. Since the number of OTUs are dependent on error rate and chimera prevalence the diversity metrics can vary simply by changing the polymerase used. Based on these observations, metrics that measure diversity using calculations such as richness depend on the polymerase, but this may not be the case for all diversity metrics.

Based on polymerase used, our observations generally found no difference when using the Bray-Curtis index to measure the bacterial community. Specifically, there was no difference in distance between successive 5-cycle increments within samples between the polymerases when using the Bray-Curtis index [Figure 3]. However, the distance between cycle comparison groups (e.g. 20 cycles versus 25 cycles group compared to 30 cycles versus 35 cycles group) was different across polymerase used. These results support the idea that cycle number and not the specific polymerase used may have a larger affect on metrics such as Bray-Curtis. One possible reason for this outcome is that many of the OTUs generated by polymerase dependent error rates and chimera prevalence are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight these OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate some of the observed polymerase dependent differences in error rate and chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996], richness) may be less robust to polymerase dependent error rates and chimera prevalence. When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. One possible way to better choose polymerases may be based on the DNA polymerase family used in the PCR mixture.

Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family because of the different binding affinities and error correction capacity [@polymerase_Ishino_2014], this is unlikely to be the only contributor. Within our study the highest and lowest chimera prevalence both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014]. Additionally, based on the information supplied by the respective manufacturers, the differences between the two PCR mixtures are not immediately apparent. Both PCR mixtures contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. Since it is not possible to know everything about the mixture beyond what was willingly provided by the manufacturer, it is possible that differences in how the recombinant *Taq* was generated or other compounds within the PCR mixture could be a contributing factor for the differences in error rate and chimera prevalence. Beyond the choice of the type of polymerase, there may be other ways to reduce the affect of polymerase dependent error rates and chimera prevalence on the downstream results.

Standardization of the 16S rRNA gene sequencing workflow may partially solve this problem by introducing a consistent bias to all samples. However, as mentioned in the introduction, many of the choices made to standardize the workflow will result in missing important members of the microbial community. Therefore, the degree of workflow standardization across studies and research groups need to be approached on a study by study basis. All aspects of the 16S rRNA gene sequencing workflow should be customized for the specific scientific questions that wants to be answered. Although a diversity of approaches may make reproducibility and replicability more difficult it will help reduce systematic biases from occuring.

### 2018-05-30 submission/manuscript.tex
complete changes to the discussion

Our observations support that the number of OTUs, error rate, and
chimera prevalence depends on polymerase used and that chimera
prevalence is highly correlated with the number of OTUs {[}Figure
1-5{]}. Our observations also suggest that undetected chimeras, not
removed using standard bioinformatic approaches, cause many of these
differences. In addition to the reported polymerase dependent
differences, the bacterial community also varied by the number of cycles
used. These differences are important because many diversity metrics
rely on calculations (e.g.~richness) that are dependent on the number of
OTUs. Since the number of OTUs are dependent on error rate and chimera
prevalence the diversity metrics can vary simply by changing the
polymerase used. Based on these observations, metrics that measure
diversity using calculations such as richness depend on the polymerase,
but this may not be the case for all diversity metrics.

There were few differences that depend on polymerase for between sample
diversity, as measured by the Bray-Curtis index. Our observations
generally found no differences in the overall bacterial community
composition for the number of cycles used. One possible reason for this
outcome is that our study did not have enough power to detect
differences due to low sample number in each group. Another reason is
that many of the OTUs are likely not highly abundant, allowing the
Bray-Curtis index to be able to successfully down-weight chimeric OTUs
(Minchin, 1987). The choice of downstream diversity metric could be an
important consideration in helping to mitigate these observed polymerase
dependent differences in chimera prevalence. Metrics that solely use
presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas, 1996)) may be
less robust to chimera prevalence and by extension total number of OTU
differences in polymerases. When choosing a distance metric, careful
consideration of the biases introduced from the PCR step of the 16S rRNA
gene sequencing workflow need to be taken into account. With differences
in the number of OTUs and chimera prevalence depending on polymerase
used, it might be easier to avoid specific DNA polymerase families
altogether.

Although the variation in error rate and chimera prevalence may be due
to the DNA polymerase family, this is unlikely to be the only
contributor. For example, the highest and lowest chimera rates both
belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the material safety data sheet the differences
between the two polymerases are not immediately apparent. Both
polymerases contain a recombinant \emph{Taq} DNA polymerase, a
\emph{Pyrococcus} spp GB-D polymerase and a platinum \emph{Taq}
antibody. With everything else being equal, it is possible that
differences in how the recombinant \emph{Taq} was generated could be a
contributing factor for the differences observed between the
polymerases. We are unlikely to avoid adding polymerase dependent bias
to 16S rRNA gene sequencing results, however, these differences may not
be large enough to mask the actual biological signal.

The majority of polymerases we studied add small increases in the number
of OTUs and chimera prevalence and may be masked by biological
differences. The sequence error introduced by the polymerase is also
small and likely to be smaller than the biological variation within a
specific study, which would be consistent with previous findings for
preservation and DNA extraction methods (Salter et al., 2014; Song et
al., 2016; Luo et al., 2016). The choice of polymerase should be an
important consideration in the creation of a 16S rRNA gene sequencing
workflow because using different polymerases can add error and bias to
the downstream observations. Although standardization of the workflow
may partially solve this problem by introducing a consistent bias to all
samples, it does come at a cost.

Methods can be standardized but they commonly contain bias that is
reproducible and may miss important associations. Bias can be easily
reproduced and can be found in every step of the 16S rRNA gene
sequencing workflow (Salter et al., 2014; Gohl et al., 2016; Luo et al.,
2016; Amir et al., 2017). This study shows that specific diversity
metrics used to measure the microbial community consistently vary based
on polymerase. Standardizing multiple workflows to one specific
polymerase could be detrimental since some polymerases may work better
in certain situations over others. Arguably, the degree of workflow
standardization across studies and research group needs to be approached
on a study by study basis and not every project needs to use the exact
same approach. All aspects of the 16S rRNA gene sequencing workflow need
to be customized for the specific microbial community that will be
sampled. Although a diversity of approaches may make reproducibility and
replicability more difficult it will help to avoid systematic biases
from occuring due to widespread standardization of approaches.

@@

\hypertarget{ref-Amir2017}{}
Amir A., McDonald D., Navas-Molina JA., Debelius J., Morton JT., Hyde
E., Robbins-Pianka A., Knight R. 2017. Correcting for microbial blooms
in fecal samples during room-temperature shipping. \emph{mSystems}
2:e00199--16. DOI:
\href{https://doi.org/10.1128/msystems.00199-16}{10.1128/msystems.00199-16}.

@@

\hypertarget{ref-preservation_Luo_2016}{}
Luo T., Srinivasan U., Ramadugu K., Shedden KA., Neiswanger K., Trumble
E., Li JJ., McNeil DW., Crout RJ., Weyant RJ., Marazita ML., Foxman B.
2016. Effects of specimen collection methodologies and storage
conditions on the short-term stability of oral microbiome taxonomy.
\emph{Applied and Environmental Microbiology} 82:5519--5529. DOI:
\href{https://doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.

>>>>

Our observations support that the number of OTUs, error rate, and
chimera prevalence depends on polymerase used and that chimera
prevalence is highly correlated with the number of OTUs {[}Figure 1-2 \&
4-6{]}. Our observations also suggest that undetected chimeras, not
removed using standard bioinformatic approaches, cause many of these
differences. In addition to the reported polymerase dependent
differences, the bacterial community also varied by the number of cycles
used. These differences are important because many diversity metrics
rely on calculations (e.g.~richness) that are dependent on the number of
OTUs. Since the number of OTUs are dependent on error rate and chimera
prevalence the diversity metrics can vary simply by changing the
polymerase used. Based on these observations, metrics that measure
diversity using calculations such as richness depend on the polymerase,
but this may not be the case for all diversity metrics.

Based on polymerase used, our observations generally found no difference
when using the Bray-Curtis index to measure the bacterial community.
Specifically, there was no difference in distance between successive
5-cycle increments within samples between the polymerases when using the
Bray-Curtis index {[}Figure 3{]}. However, the distance between cycle
comparison groups (e.g.~20 cycles versus 25 cycles group compared to 30
cycles versus 35 cycles group) was different across polymerase used.
These results support the idea that cycle number and not the specific
polymerase used may have a larger affect on metrics such as Bray-Curtis.
One possible reason for this outcome is that many of the OTUs generated
by polymerase dependent error rates and chimera prevalence are likely
not highly abundant, allowing the Bray-Curtis index to be able to
successfully down-weight these OTUs (Minchin, 1987). The choice of
downstream diversity metric could be an important consideration in
helping to mitigate some of the observed polymerase dependent
differences in error rate and chimera prevalence. Metrics that solely
use presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas, 1996),
richness) may be less robust to polymerase dependent error rates and
chimera prevalence. When choosing a distance metric, careful
consideration of the biases introduced from the PCR step of the 16S rRNA
gene sequencing workflow need to be taken into account. One possible way
to better choose polymerases may be based on the DNA polymerase family
used in the PCR mixture.

Although the variation in error rate and chimera prevalence may be due
to the DNA polymerase family because of the different binding affinities
and error correction capacity (Ishino \& Ishino, 2014), this is unlikely
to be the only contributor. Within our study the highest and lowest
chimera prevalence both belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the information supplied by the respective
manufacturers, the differences between the two PCR mixtures are not
immediately apparent. Both PCR mixtures contain a recombinant \emph{Taq}
DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
\emph{Taq} antibody. Since it is not possible to know everything about
the mixture beyond what was willingly provided by the manufacturer, it
is possible that differences in how the recombinant \emph{Taq} was
generated or other compounds within the PCR mixture could be a
contributing factor for the differences in error rate and chimera
prevalence. Beyond the choice of the type of polymerase, there may be
other ways to reduce the affect of polymerase dependent error rates and
chimera prevalence on the downstream results.

Standardization of the 16S rRNA gene sequencing workflow may partially
solve this problem by introducing a consistent bias to all samples.
However, as mentioned in the introduction, many of the choices made to
standardize the workflow will result in missing important members of the
microbial community. Therefore, the degree of workflow standardization
across studies and research groups need to be approached on a study by
study basis. All aspects of the 16S rRNA gene sequencing workflow should
be customized for the specific scientific questions that wants to be
answered. Although a diversity of approaches may make reproducibility
and replicability more difficult it will help reduce systematic biases
from occuring.

@@

@@
